---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)

```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class
-   Sex
-   Age
-   Survived
-   n?

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   2224 according to wikipedia
-   Are there any differences?
    -   23 passanger difference
-   If yes, what might account for those differences?
    -   People that were never found, confirmed dead or alive
    -   People that weren't well documented / they didn't have the names of
    -   people that they didn't know the class/age or sex of

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  ggplot(aes(x = Sex, y=n, fill = Class)) +
  geom_col(position = "dodge")+
  facet_wrap(~Survived)
```

**Observations**:

-   Of the people that survived it was nearly 50/50 male female, but almost all of the people that died were male
-   The largest surviving group was male crew
-   Many of the people who died were crew
-   1st class females were the least likely to die
-   of females the most who died where 3rd class
-   despite more crew overall seeming to die compared to 3rd class. female crew almost all survived while many 3rd class females died

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop%>%
  ggplot(aes(x = Sex, y=Prop, fill = Class)) +
  geom_col(position = "dodge")+
  facet_wrap(~Survived)
```

**Observations**:

-   despite having the highest number of survivors male crew actually have the lowest proportion of survival
-   Female crew have one of the highest proportions of survival in comparison
-   1st class woman have almost no deaths
-   Is there anything *fishy* going on in your plot?
    -   We haven't seperated by age, so this allows for totals that are over 100% for some categories such as 1st and 2nd class men and woman. For whatever reason this is maxing them at 100% and not summing it past that.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop%>%
  ggplot(aes(x = Sex, y=Prop, fill = Class)) +
  geom_col(position = "dodge")+
  facet_grid(Age~Survived)
```

**Observations**:

-   1st/2nd class children where the highest priority, followed by woman.
-   Woman in the crew were highly valued second only to first class woman
-   For some reason more 3rd class men survived then 2nd class men proportionally
    -   This might have to do with there being far more 3rd class men in general,
    -   or perhaps the location of the 2nd class rooms relative to the life boat
-   The only children who died where 3rd class
-   Unlike other females, 3rd class females only had a 50% chance of survival
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   The children in 1st/2nd class all survived, however with this new graph we can now see that the adults did not all survive, particularly the men. In fact the vast majority of the 2nd class men didn't survive, and only about 30% of the woman.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
