---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
##glimpse(gapminder)
head(gapminder)
summary(gapminder)
```

**Observations**:

-   Country, Continent, Year, Life expectancy, Population

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
    gapminder %>%
      pull(year) %>%
        max()
year_min <- gapminder %>%
      pull(year) %>%
        min()

print(year_max)
print(year_min)
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
year_min_data <-
  filter(gapminder,year==1952)
print(year_min_data)
year_min_data %>%
  ggplot(aes(continent, gdpPercap)) +
    geom_point(size = 2, alpha = .2) +
    theme_minimal() + 
    scale_y_continuous(trans='log10')
  
```

**Observations**:

-   Americas and europe have the highest max GDPs
-   Africa has the lowest average GDP
-   Asia has a very high outlier
-   Europe has a pretty consistent tight set of data but theres an strong outlier on both the high and low ends\

**Difficulties & Approaches**:

-   The outlier in Asia makes it very different to distinguish the rest of the data since its such a higher value than anything else. Plus theres a significant number of countries with almost no GDP
    -   Used a log scale to make the data more readable
-   Overlapping points makes it difficult to estimate the density of the data
    -   set alpha to .2

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
##Outlier 1, asia
Asia_outlier <-
  gapminder %>%
    filter(year==1952) %>%
      filter(continent=="Asia") %>%
        arrange(desc(gdpPercap))
print(Asia_outlier)
##outlier 2/3 europe
Asia_outlier <-
  gapminder %>%
    filter(year==1952) %>%
      filter(continent=="Europe") %>%
        arrange((gdpPercap))
print(Asia_outlier)
Asia_outlier <-
  gapminder %>%
    filter(year==1952) %>%
      filter(continent=="Europe") %>%
        arrange(desc(gdpPercap))
print(Asia_outlier)
```

**Observations**:

-   Identify the outlier countries from q2
    -   High Outlier in Asia is Kuwait
    -   High in Europe is Switzerland
    -   Low in Europe is Bosnia

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
year_min_max_data <-
  filter(gapminder, year==max(year) | year==min(year))
print(year_min_max_data)
year_min_max_data %>%
  ggplot(aes(continent, gdpPercap)) +
    geom_point(size = 2, alpha = .2, mapping=aes(color=factor(year))) +
    theme_minimal() + 
    scale_y_continuous(trans='log10')+
    geom_point(
      data = . %>% filter(country %in% c("Kuwait", "Switzerland", "Bosnia and Herzegovina")),
      mapping = aes(color = factor(year),shape=country),
      size = 4
  )+
    scale_shape_manual(values = c("Kuwait" = 2, "Switzerland" = 3, "Bosnia and Herzegovina" = 4 ))
  
```

**Observations**:

-   Kuwait has remained as the highest GDP in asia as well as one of the highest in the whole world but is no longer a nearly as significant of an outlier
-   Switzerland has lost its lead on europe and and while still near the top is no longer the highest or an outlier in the same way
-   Bosnia remains near the bottom of europe but has risen in GDP with the rest of the rejoin as is no longer the lowest GDP in the region.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
data_summary <- gapminder %>%
  group_by(country,continent) %>%
  summarize(
    avg_lifeExp = mean(lifeExp, na.rm = TRUE),
    avg_pop = mean(pop, na.rm = TRUE),
    avg_gdpPercap = mean(gdpPercap, na.rm = TRUE),
    .groups = "drop"  # optional: ungroup the result
  )
data_summary <-
  data_summary %>%
    mutate(percent_gdp=100*(avg_gdpPercap/(sum(avg_gdpPercap))))
print(data_summary)
data_summary %>%
  ggplot(aes(avg_gdpPercap)) +
      geom_histogram(bins=50) +
      theme_minimal()
      #scale_y_continuous(trans='log10')
data_summary %>%
  filter(avg_gdpPercap<30000) %>%
    ggplot(aes(avg_gdpPercap)) +
      geom_histogram(bins=100) +
      theme_minimal()
data_summary %>%
  ggplot(aes(avg_gdpPercap,percent_gdp,color=continent)) +
      geom_col() +
      theme_minimal()
      #scale_x_continuous(trans='log10')
data_summary %>%
  ggplot(aes(avg_gdpPercap,percent_gdp,color=continent)) +
      geom_col() +
      theme_minimal() +
      scale_x_continuous(trans='log10')

```

-   Question:
    -   Distribution of average GDPs over the timespan
-   This graph shows the inequity in our economic systems. There are far more countries with very low GDPs than there are countries in the middle of the gdp range. The non log graph shows even more strongly how some countries control far more money than many poorer countries put together.
-   This inspired my next set off graphs that look at percentage of overall GDP global based on average gdp
    -   the largest gdp country in Asia controls the largest percent of global gdp at over 6 percent, meanwhile many of the countries in Asia have an almost immeasurable percentage of global gdp
    -   in Europe many countries have between 2-3 percent of global gdp
    -   logically, this graph shows a directly linear relationship

```{r q5-task2}
## TASK: Your second graph
data_summary %>%
  ggplot(aes(avg_pop,avg_gdpPercap, color=continent)) +
      geom_point() +
      theme_minimal() + 
      scale_x_continuous(trans='log10')
data_summary %>%
  ggplot(aes(avg_pop,avg_gdpPercap, color=continent)) +
      geom_col() +
      theme_minimal() + 
      scale_x_continuous(trans='log10')
```

-   (Your notes and observations here)
-   Question: How does population effect GDP?
    -   Highest populations are in Asia
    -   Both very high population and very low population seems to lead to a lower average GDP
    -   In general there doesn't seem to be a strong correlation between avg population and avg gdp

```{r q5-task3}
## TASK: Your third graph
# 1. Calculate the average GDP per country (across all years). Averaging first so that a country can't change categories over the course of time
country_gdp_avg <- 
  gapminder %>%
    group_by(country) %>%
    summarise(
      avg_gdpPercap = mean(gdpPercap, na.rm = TRUE),
      .groups = "drop"
    )

# 2. Bin countries based on their average GDP
country_gdp_avg <- 
  country_gdp_avg %>%
    mutate(
      gdp_bin = cut(
        avg_gdpPercap,
        breaks = quantile(avg_gdpPercap, probs = seq(0, 1, by = 0.2), na.rm = TRUE),
        include.lowest = TRUE,
        labels = c("Very Low", "Low", "Medium", "High", "Very High")
      )
    )

# 3. Merge the GDP bin info back into the full dataset
df_with_bins <- 
  gapminder %>%
    left_join(country_gdp_avg %>% select(country, gdp_bin), by = "country")

# 4. Group by year and gdp_bin, then average other values
df_summary <- 
  df_with_bins %>%
    group_by(year, gdp_bin, continent) %>%
    summarise(
      avg_lifeExp = mean(lifeExp, na.rm = TRUE),
      avg_pop = mean(pop, na.rm = TRUE),
      avg_gdpPercap = mean(gdpPercap, na.rm = TRUE),
      .groups = "drop"
    )
print(df_with_bins)
print(df_summary)

df_summary  %>%
  ggplot(aes(year,avg_pop, color=gdp_bin)) +
      geom_line() +
      theme_minimal() +
      facet_wrap(~continent) +
      scale_y_continuous(trans='log10')
df_with_bins %>%
  ggplot(aes(year,pop, color=gdp_bin,group=country)) +
    geom_line() +
    theme_minimal() +
    facet_grid(continent~gdp_bin) +
    scale_y_continuous(trans='log10') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

-   How does average GDP range of a country effect population growth of that country over time, and is it effected by region?
    -   **Cool thing**
        -   At first I binned the entire dataset, however this messed up my graphs as some countries moved between bins overtime so instead I averaged the countries gdp first before sorting it into a bin
    -   I used a log scale to make population changes more apparent
    -   Africa saw the most consistent steady population growth across gdp bins, with gdp having little effect on population growth.
    -   Europes population growth is relatively flat for high and very high gdps, but there is significant growth for medium gpd countries
    -   High gdp in asia has a higher growth than very high gdp
-   I was worried that individual countries was throwing of the results I was seeing in the bins so I also made a facet_grid that plots lines for all the countries at once over time, still visualized by gdp bin, but not averaged
    -   This gave a much more detailed view into the data
        -   For example two of the 3 countries in the medium gdp bin for europe had significant growth, but the third did not
        -   Some high income countries in Asia are have very large population changes while most are relatively flat over time
        -   Medium income countries in the Americas follow a very consistent curve over time across countries with limited variance.\
