---
title: "Antibiotics"
author: "(Your name here)"
date: 2020-
output:
  github_document
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective. The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*. A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
glimpse(df_antibiotics)
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`. These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long. You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
#Lets start off easy with a lineplot
data_long <- df_antibiotics %>%
  pivot_longer(cols = c("penicillin","streptomycin","neomycin"), 
               names_to = "Antibiotic", 
               values_to = "Mic")
print(data_long)
data_long %>%
   ggplot(aes(Antibiotic,Mic,group=bacteria, color=bacteria)) +
      geom_line(aes(linetype = gram)) +
      geom_point()  +
      theme_minimal()+
      scale_y_continuous(trans='log10')+
      geom_abline(aes(intercept = log10(1E-01), slope=0), col="red", lwd=.5)
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
data_long %>%
   ggplot(aes(Antibiotic,Mic, color=gram)) +
      geom_col() +
      facet_grid(~bacteria) +
      theme_minimal()+
      geom_abline(aes(intercept = log10(1E-01), slope=0), col="red", lwd=.5)+
      scale_y_continuous(trans='log10')+
      theme(axis.text.x = element_text(angle = 80, hjust = 1))+
      theme(strip.text = element_text(angle = 88))
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
#is there a correlation between two effectivenesses 
df_antibiotics %>%
   ggplot(aes(penicillin,neomycin, color=bacteria)) +
      geom_point()  +
      theme_minimal()+
      scale_y_continuous(trans='log10')+
      scale_x_continuous(trans='log10')

```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
# WRITE YOUR CODE  Box Plot
data_long %>%
   ggplot(aes(bacteria, Mic, color=gram)) +
      geom_boxplot()  +
      theme_minimal()+
      scale_y_continuous(trans='log10') +
      theme(axis.text.x = element_text(angle = 80, hjust = 1))+
      geom_abline(aes(intercept = log10(1E-01), slope=0), col="red", lwd=.5)
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE Heat Map
data_long %>%
   ggplot(aes(bacteria, Antibiotic, fill=Mic, color=gram)) +
      geom_tile()  +
      theme_minimal() +
      geom_text(aes(label = ifelse(Mic<=.1, "+", "")), color = "black", size = 3) +
      geom_point(data = . %>% filter(Mic<=.1),
               aes(bacteria, Antibiotic, shape = "Mic Under .1"),
               color = "black", size = 3, inherit.aes = FALSE) +
      scale_shape_manual(name = "", values = c("Mic Under .1" = 3)) +
      theme(
      axis.text.x = element_text(angle = 80, hjust = 1),
      legend.key.size = unit(0.5, "lines"),    # smaller legend boxes
      legend.text = element_text(size = 8)     # smaller text
      )+
      scale_fill_continuous(trans='log10') +
      theme(axis.text.x = element_text(angle = 80, hjust = 1))
      
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions. Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below. You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions. **Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

*Observations* -\
What is your response to the question above? - Penicillin seems to be either the most effective or the least effective at fighting bacteria. Neomycin is the most effective general antibiotic against a wide range of bacteria. Neomycin is also the best antibiotic against gram negative bacteria with penecilin not being effective against any and streptomycin only being effective on 1 gram negative bacteria. Streptomycin isn't effective against any unique bacteria that one of the others isn't effective against.

\
Which of your visuals above (1 through 5) is **most effective** at helping to answer this question? - Graphs 1 and 5 are most effective.

Why? - This question benefits from a graph that shows as much information at once which graphs 1 and 5 excel at. Additionally, they both clearly identify which values fall under the threshold of .1 which is very advantageous for identifying effectiveness! Personally I find the colors in graph 5 more helpful than the dashed lines in graph 1 for looking at gram positive versus negative. However in terms of relative performance between antibiotics on the same bacteria graph 1 is much more visually clear than the heatmap. Graph 4 doesn't identify specific antibiotics, graph 3 is more specific than this question, and graph 2 is harder to pull summary information from due to the size and spread of the data on the box plots.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

*Observations* - What is your response to the question above? Diplococcus Pneumoniae was renaimed to Streptococcus Pneumoniae because like the other Streptococcus bacteria penicillin is very effective on Diplococcus Pneumoniae while Neomycin is very ineffective on the bacteria. This can be seen in the top left corner of graph 3. On the other hand, *Streptococcus fecalis* sees poor response to penicillin.

\- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question? - Graph 2 is the most effective

\- Why? - Graph 2 clearly groups bacteria into groups based on the comparative effectiveness of Penicillin and Neomycin, this is perfect for noticing the trends between the Streptococcus bacteria.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
